Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package

### 2021-06-21 _book/_main.tex
Revise for JOSS submission. Address Ed's comments.

\author{
\\
}
\title{Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package}

\Plainauthor{}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}

\Abstract{
It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the \textbackslash pkg\{Bigsimr\} R package. This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix. \textbackslash pkg\{Bigsimr\} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to \(d=10,000\). Finally, we demonstrate example applications enabled via \textbackslash pkg\{Bigsimr\} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
}

@@

% Pandoc header
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}

\usepackage{setspace}
\setstretch{2.0}

@@

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Chernick2008} to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings \citep{MB13, XZ19}. For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with flexible, robust software available \citep{Yan2007, Chen2001}. Yet these approaches do not scale in a timely fashion to high-dimensional problems \citep{Li2019gpu}. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see \citet{Park1996}). While more recent approaches \citep{MB13, Xia17, BF17} have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\). Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, provide Pearson matching with some caveats.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. Further, we introduce the \texttt{bigsimr} R package that provides high-performance software implementing our algorithm. The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.

Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. Next, we detail an illustrative low-dimensional example of basic use of the \texttt{bigsimr} R package. Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy. We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

@@

The \texttt{bigsimr} \texttt{R} package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i\).

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and ``scalability'' properties (SP). The BPs are adapted from an existing criteria due to \citet{Nik13a}. A suitable simulation strategy should possess the following properties:

@@

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
\end{itemize}

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes \citep{BE07, Schissler2018}.

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2020-02-25)
R> num_genes    <- 3
R> num_patients <- 5
R> gene_sample  <- sample(example_genes[1:1000], num_genes)
R> 
R> cap <- paste0("mRNA expression for three selected high-expressing genes, ",
+               paste(gene_sample, collapse = ", "), 
+               ", for the first five patients in the TCGA BRCA data set.")
R> 
R> small_brca <- example_brca %>%
+   select(all_of(gene_sample)) %>%
+   head(n = num_patients) %>%
+   as_tibble(rownames = "Patient ID") %>%
+   mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
R> 
R> small_brca %>%
+   kable(format = "latex", booktabs = TRUE, caption = cap)
\end{CodeInput}
\begin{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> ggpairs(
+   data = example_brca[, gene_sample],
+   upper = list(continuous = wrap('cor', method = "spearman"))
+ ) + 
+   theme_bw() +
+   theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
\end{CodeInput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{_main_files/figure-latex/ch010-realDataFig-1}

@@

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}. Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby captures nonlinear yet monotone relationships. Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance meaning that orderings have the same direction (e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of the values, not the values themselves.

Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates. As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

@@

There is a strong connection of our simulation strategy to Gaussian \emph{copulas} (see \citet{Nelsen2007} for a technical introduction). A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability structures. Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
\label{eq:copula}
\end{equation}

\(x_i \in {\mathbb R}, i=1,\ldots,d.\)

A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\). This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \({\bf R_P}\). As the marginals are standardized to have unit variance, however, \({\bf R_P}\) is a Pearson correlation matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through

\begin{equation}
\label{eq:gauss}
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where \(\Phi(\cdot)\) is the standard normal CDF. Note that the copula \(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.~\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

@@

This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\). The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a high-dimensional NORTA-inspired algorithm.

@@

where \(\phi(\cdot)\) is the joint probability distribution function of two correlated standard normal variables. For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see \citep{Xia17} and the publicly available \texttt{nortaRA} R package \citep{Chen2001}. As described later, to enable HD Pearson matching we approximate these integrals.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated. Even in low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed \emph{NORTA defective}.

\citet{GH02} conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\). They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as \citet{GH02} also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\). This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

@@

We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \tightlist
    \item
      Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
    \end{enumerate}
  \item
    (i') Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    (i'\,') Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check admissibility and, if needed, compute the nearest correlation matrix.

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \tightlist
    \item
      Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
    \end{enumerate}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \setcounter{enumii}{1}
    \tightlist
    \item
      If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise
    \end{enumerate}
  \item
    (ii') Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
  \end{itemize}
\item
  Gaussian copula

>>>>

\author{
Alfred G. Schissler\\University of Nevada \And Edward J. Bedrick\\University of Arizona \And Alexander D. Knudson\\University of Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin Nguyen\\University of Nevada \And Anna K. Panorska\\University of Nevada \AND Juli Petereit\\University of Nevada \And Walter W. Piegorsch\\University of Arizona \And Duc Tran\\University of Nevada
}
\title{Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package}

\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D. Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli Petereit, Walter W. Piegorsch, Duc Tran}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data}

\Abstract{
It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. To meet this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces. These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix. \pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to \(d=10,000\). We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
}

@@

% Pandoc header
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{setspace}

@@

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Chernick2008} to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings \citep{MB13, XZ19}. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with existing software available \citep{Yan2007, Chen2001}. Yet these approaches do not scale in a timely fashion to HD problems \citep{Li2019gpu}. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see \citet{Park1996}). While more recent approaches \citep{MB13, Xia17, BF17} have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\). Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution. For our method, we developed a high-performance implementation: the \texttt{Bigsimr} Julia package, with R and Python interfaces \texttt{bigsimr}.

This article proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. We proceed by providing an illustrative LD \texttt{bigsimr} workflow. Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

@@

The \texttt{Bigsimr} package presented here provides multiple high-performance algorithms that operate with HD multivariate data. All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i, i=1,\ldots,d\).

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and `scalability' properties (SP). The BPs are adapted from an existing criteria due to \citet{Nik13a}. A suitable simulation strategy should possess the following properties:

@@

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.
\end{itemize}

@@

Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes \citep{BE07, Schissler2018}.

We illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments). For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.

\begin{CodeChunk}
\begin{table}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch010-realDataFig-1}

@@

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}. Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby capturing nonlinear yet monotone relationships. Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning that orderings have the same direction, e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of the values, not the values themselves. Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates. As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

@@

There is a connection of our simulation strategy to Gaussian \emph{copulas} \citep{Nelsen2007}. A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability. Consequently, the study of copulas is an important and active area of statistical theory and practice. For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
\label{eq:copula}
\end{equation}

A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\), where \(\Phi(\cdot)\) is the standard normal CDF. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \(R_P\). Since the marginals are standardized to have unit variance, however, \(R_P\) is a Pearson correlation matrix. If \(F_R\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_R\) is defined through

\begin{equation}
\label{eq:gauss}
F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

Here the copula function \(C_R\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).

Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz. \(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).

\hypertarget{other-multivariate-simulation-packages}{%
\subsection{Other Multivariate Simulation Packages}\label{other-multivariate-simulation-packages}}

To place \texttt{Bigsimr}'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages. In \texttt{R}, the \texttt{MASS} package \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks. Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}. For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}. High-dimesional random vectors via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}. The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence modeling. Table \ref{tab:compare-table} summarizes the above discussion.

\begin{table}[h]
\centering
\caption{\label{tab:compare-table}Comparison of selected features of \pkg{Bigsimr} to existing packages for multivariate simulation.}
\begin{tabular}{@{}llll@{}}
Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

\clearpage

@@

This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\). The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a HD NORTA-inspired algorithm.

@@

where \(\phi(\cdot)\) is the joint probability distribution function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite polynomial approximation of these \(\binom{d}{2}\) double integrals, as suggested by \citet{XZ19}.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed \emph{NORTA defective}.

\citet{GH02} conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\). They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as \citet{GH02} also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\). This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

@@

We now describe our algorithm to generate random vectors, which extends NORTA to HD via nearest correlation matrix replacement and provides rank-based dependency matching:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step, either

  \begin{itemize}
  \tightlist
  \item
    Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check admissibility and, if needed, compute the nearest correlation matrix.

  \begin{itemize}
  \tightlist
  \item
    Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
  \item
    If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise,
  \item
    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
  \end{itemize}
\item
  Gaussian copula

  \begin{itemize}
  \tightlist
  \item
    Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
  \item
    Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz. \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
  \item
    Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
  \end{itemize}
\end{enumerate}

\emph{Step 1: Mapping step.} We first employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert()}). Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
